\pkg{SWMPr}: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries

### 2015-09-02 swmpr_manu.tex
text butchered... need to seriously shorten

\subsection{Installing the package}

The \pkg{SWMPr} package was developed for use with the R ($\geq$ v3.0.0) statistical programming language. The \pkg{SWMPr} package can be downloaded from CRAN by executing the following commands at the R console prompt.  The package is loaded in the workspace using the \code{library} command.

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> install.packages('SWMPr')
> library(SWMPr)
\end{verbatim}
\end{kframe}
\end{knitrout}

The \pkg{SWMPr} package was developed by considering a common data workflow that categorizes the functions as one of three steps based on their intended use: \Bigtxt{retrieving}, \Bigtxt{organizing}, and \Bigtxt{analyzing}.  Functions for retrieving are used to import the data into R as a \code{"swmpr"} object class.  Functions for organizing and analyzing the data provide methods for working with a \code{"swmpr"} object.  An additional group of `miscellaneous' functions are included as helpers for the main functions.  The following describes the package structure, beginning with the retrieval functions, a description of the \code{"swmpr"} object returned after retrieval, and, finally, the organizing and analyzing functions.

@@

Data retrieval functions to import data directly into R from the CDMO include \code{all\_params}, \code{all\_params\_dtrng}, and \code{single\_param}: \code{all\_params} returns the most recent records of all parameters at a station, \code{all\_params\_dtrng} returns all records within a date range for all parameters or a single parameter, and \code{single\_param} is identical to \code{all\_params} except that a single parameter is requested. Due to rate limitations on the CDMO server, the retrieval functions return a limited number of records with each request.  However, the \pkg{SWMPr} functions use the native CDMO web services iteratively (i.e., within a loop) to obtain all requested records.  Download time can be excessive for longer time series.

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # all parameters for a station, most recent
> all_params('hudscwq')
> 
> # get all parameters within a date range
> all_params_dtrng('hudscwq', c('09/10/2012', '02/8/2013'))
> 
> # get single parameter within a date range
> all_params_dtrng('hudscwq', c('09/10/2012', '02/8/2013'), 
+   param = 'do_mgl')
> 
> # single parameter for a station, most recent
> single_param('hudscwq', 'do_mgl')
\end{verbatim}
\end{kframe}
\end{knitrout}

The second approach for data retrieval is to use the \code{import\_local} function to import data into R that are locally available after downloading from CDMO.  This approach is most appropriate for large, specific data requests. The \code{import\_local} function is designed for data from the zip downloads feature in the advanced query section of the CDMO website. The zip downloads feature can be used to obtain data from multiple stations in one request.  The downloaded data will be in a compressed folder that includes multiple .csv files by year for a given data type (e.g., apacpwq2002.csv, apacpwq2003.csv, apacpnut2002.csv, etc.).  The \code{import\_local} function can be used to import files directly from the zipped folder or after the folder is decompressed.

@@

The \code{aggreswmp} function aggregates parameter data for a \code{"swmpr"} object by set units of time.  This function is most useful for aggregating noisy data to evaluate trends on longer time scales or to simply reduce the size of a dataset.  Data can be aggregated by years, quarters, months, weeks, days, or hours by a user-defined function, which defaults to the mean.  A \code{"swmpr"} object is returned for the aggregated data, although the \code{datetimestamp} vector will be converted to a date object if the aggregation period is a day or longer.  Days are assigned to the date vector if the aggregation period is a week or longer based on the \code{round} method for IDate objects created in the \CRANpkg{data.table} package \citep{Dowle14}.  Additionally, the method of treating \code{NA} values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned, particularly for nutrient data (see the example below).

% table for analysis functions
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Analysis functions available from the \\pkg{SWMPr} package.  Full documentation for each function is in the help file (e.g., execute \\code{?aggreswmp} for individual functions or \\code{help.search(`analyze', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:analyze",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Analysis functions available from the \pkg{SWMPr} package.  Full documentation for each function is in the help file (e.g., execute \code{?aggreswmp} for individual functions or \code{help.search(`analyze', package = `SWMPr')} for all).\label{tab:analyze}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\code{aggreswmp}&Aggregate \code{"swmpr"} objects for different time periods - years, quarters, months,  weeks, days, or hours.  The aggregation function defaults to the mean.\tabularnewline
\code{aggremetab}&Aggregate metabolism data from a \code{"swmpr"} object.  This is primarly used within \code{plot\_metab} but may be useful for simple summaries of daily metabolism data.\tabularnewline
\code{ecometab}&Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method.\tabularnewline
\code{decomp}&Decompose a \code{"swmpr"} time series into trend, seasonal, and residual components.  This is a simple wrapper to \code{decompose} \citep{Kendall83}.  Decomposition of monthly or daily trends is possible.\tabularnewline
\code{decomp\_cj}&Decompose a \code{"swmpr"} time series into grandmean, annual, seasonal, and events components.  This is a simple wrapper to \code{decompTs} in the \CRANpkg{wq} package \citep{Jassby14}.  Only monthly decomposition is possible.\tabularnewline
\code{hist}&Plot a histogram for a \code{"swmpr"} object.\tabularnewline
\code{lines}&Add lines to an existing plot created with \code{plot}.\tabularnewline
\code{map\_reserve}&Create a map of all stations in a reserve using the \CRANpkg{ggmap} package \citep{Kahle13}.\tabularnewline
\code{na.approx}&Linearly interpolate missing data (\code{NA} values) in a \code{"swmpr"} object. The maximum gap size that is interpolated is defined by the arguments.\tabularnewline
\code{plot}&Plot a univariate  time series for a \code{"swmpr"} object.  The parameter name must be specified.\tabularnewline
\code{plot\_metab}&Plot ecosystem metabolism estimates after running \code{ecometab} on a combined \code{"swmpr"} object.\tabularnewline
\code{plot\_summary}&Create summary plots of seasonal/annual trends and anomalies for a water a single paramter of interest.\tabularnewline
\code{smoother}&Smooth \code{"swmpr"} objects with a moving window average.  Window size and sides (e.g., centered) can be specified, passed to \code{filter}.\tabularnewline
\hline
\end{tabular}\end{center}

@@

{\centering \includegraphics[width=0.8\textwidth]{figure/smooth_ex-1}

@@

{\centering \includegraphics[width=0.8\textwidth]{figure/interp_ex-1}

@@

{\centering \includegraphics[width=0.8\textwidth]{figure/decomp_ex1-1}

@@

{\centering \includegraphics[width=0.8\textwidth]{figure/decomp_ex2-1}

@@

{\centering \includegraphics[width=0.8\textwidth]{figure/summary_ex-1}

@@

Finally, estimates of ecosystem metabolism provide a measure of overall system productivity to evaluate whether an ecosystem is a net source or sink of organic material.  The open-water method \citep{Odum56} is a common approach to quantify metabolism using a mass balance equation that describes the change in dissolved oxygen over time from the balance between photosynthetic and respiration processes, corrected using an empirically constrained air-sea gas diffusion model \citep{Ro06,Thebault08}. The diffusion-corrected dissolved oxygen (DO) flux estimates are averaged separately over each day and night of the time series. The nighttime average DO flux is used to estimate respiration rates, while the daytime DO flux is used to estimate net primary production. To generate daily integrated rates, respiration rates are assumed constant such that hourly night time DO flux rates are multiplied by 24. Similarly, the daytime DO flux rates are multiplied by the number of daylight hours, which varies with location and time of year, to yield net daytime primary production. Respiration rates are subtracted from daily net production estimates to yield gross production rates. The metabolic day is considered the 24 hour period between sunsets on two adjacent calendar days

The \code{ecometab} function is used to implement the open-water method with a combined water quality and weather dataset \citep{Caffrey14}.  Several assumptions must be met for a valid interpretation of the results.  First, the DO time series is assumed to be a sample of the same water mass over time.  Tidal advection may have a significant influence on the time series, which can contribute to a substantial amount of noise in metabolic estimates.  The extent to which tidal advection influences the dissolved oxygen signal depends on various site-level characteristics and an intimate knowledge of the site may be required.  Second, areal rates for gross production and total respiration are based on volumetric rates normalized to the depth of the water column at the sampling location, which is assumed to be well-mixed, such that the water quality sensor is reflecting the integrated processes in the entire water column (including the benthos).  Water column depth is calculated as the mean value of the depth variable across the time series in the \code{"swmpr"} object. Depth values are floored at one meter for very shallow stations and 0.5 meters is also added to reflect the practice of placing sensors slightly off of the bottom.  Third, the air-sea gas exchange model is calibrated with wind data either collected at, or adjusted to, wind speed at 10 m above the surface.  The metadata should be consulted for exact height. Other assumptions may apply and relevant resources should be consulted to ensure appropriate application of the open-water method (see \citet{Kemp12,Needoba12}).

The following is an example that shows use of the \code{ecometab} function from a combined water quality and weather data set.  Monthly aggregations of the raw, daily estimates are plotted using \code{plot\_metab} (Figure~\ref{fig:metab_ex}).

@@

{\centering \includegraphics[width=0.8\textwidth]{figure/metab_ex-1}

@@

{\centering \includegraphics[width=0.8\textwidth]{figure/map_ex-1}

@@

\subsection{Miscellaneous functions}

Several additional functions are provided that do not fit the above categories (Table~\ref{tab:misc}).  These functions are generally used within the main functions but may be useful for more customized evaluation of SWMP data.

% table for misc functions
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Miscellaneous functions available from the \\pkg{SWMPr} package.  Most are used within the main functions above but may be useful for customized evaluations of SWMP data.  Full documentation for each function is in the help file (e.g., execute \\code{?calckl} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:misc",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Miscellaneous functions available from the \pkg{SWMPr} package.  Most are used within the main functions above but may be useful for customized evaluations of SWMP data.  Full documentation for each function is in the help file (e.g., execute \code{?calckl} at the command line).\label{tab:misc}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\code{calckl}&Estimate the reaeration coefficient for air-sea gas exchange.  Used in the \code{ecometab} function.\tabularnewline
\code{metab\_day}&Identify the metabolic day for each approximate 24 period in an hourly time series.  Used in the \code{ecometab} function.\tabularnewline
\code{param\_names}&Returns column names as a list for the parameter types (nutrients, weather, or water quality).  Includes QAQC columns with \code{f\_} prefix. Used in the data retrieval functions.\tabularnewline
\code{parser}&Parses HTML returned from CDMO data requests.  Used in the retrieval functions.\tabularnewline
\code{swmpr}&Creates a \code{"swmpr"} object class.  Used in the data retrieval functions.\tabularnewline
\code{time\_vec}&Converts time vectors to \code{POSIXct} objects with the appropriate time zone for a site.  Used in the data retrieval functions.\tabularnewline
\hline
\end{tabular}\end{center}

@@

This section describes three examples using the \pkg{SWMPr} package to illustrate the improved ability to synthesize and evaluate multi-year time series of estuarine data. First, the open-water method for estimating metabolism was applied to nearly all co-located water quality and weather sites at each NERRS reserve using all years of available data.  The results are provided primarily to illustrate ease of use of the functions and secondarily to provide an update on metabolism estimates using the most recent SWMP data.  \citet{Caffrey03} and \citet{Caffrey04} applied the open-water method to estimate ecosystem metabolism using five years of water quality observations at two sites at each of the NERRS reserves.  The air-sea gas exchange model also assumed a constant value for the reaeration coefficient, whereas the current \code{ecometab} function provides a more accurate estimate by including weather data in the calculation (see \citet{Caffrey14} for details).

Water quality and weather observations from January 1995 to December 2014 for all NERRS sites were obtained through a bulk data request using the zip downloads feature of CDMO.  All csv files for each station were imported into R using the \code{import\_local} function, processed using the \code{setstep} and \code{qaqc} functions, then saved locally as binary RData files. This resulted in a single \code{"swmpr"} object for each parameter at each site.  All files were then uploaded to a remote server for online access.  An additional R script was executed that retrieved and processed water quality and weather data at each reserve to estimate metabolism.  Two water quality sites with the longest time series at each reserve were used.  Mean annual metabolism values at each site, organized by region, are shown in Figure~\ref{fig:metab_plo}, whereas decadal comparisons are shown in Table~\ref{tab:metab_tab}.  All sites were generally net heterotrophic across the range of observations (i.e., sink of organic matter, in agreement with \citet{Caffrey03}), although differences were observed in early (i.e., 1995-2004) as compared to recent (2005-2014) time periods. Overall, the results indicate that between-region and within-site differences in metabolism are apparent and varying by time period, such that a more comprehensive evaluation of factors that influence metabolic rates is needed.  More importantly, use of the data retrieval, synthesis, and analysis functions to create the results illustrates the utility provided by the \pkg{SWMPr} package.

% metabolism figure
\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{figure}[!h]

{\centering \includegraphics[width=0.8\textwidth]{figure/metab_plo-1}

\caption[Aggregated estimates of net metabolism, gross production, and total respiration for two sites at each NERRS reserve]{Aggregated estimates of net metabolism, gross production, and total respiration for two sites at each NERRS reserve.  Values are daily integrated estimates as mean annual values averaged across all years with 95\% confidence intervals.  Two sites were chosen from each reserve that had the longest available time series. Sites were assigned to regions based on approximate geographic coordinates.}\label{fig:metab_plo}
\end{figure}

% table summary of trends
%latex.default(metabs_tab[, !names(metabs_tab) %in% c("site",     "region")], file = "", caption = cap, cgroup = c("NEM\\textsuperscript{a}",     "Pg", "Rt"), n.cgroup = c(2, 2, 2), colheads = rep(c("1995-2004",     "2005-2014"), 3), rowlabel = "Site", rowname = metabs_tab$site,     digits = 2, insert.bottom = foots, rgroup = unique(metabs_tab$region),     n.rgroup = nrgroups, label = "tab:metab_tab", size = "footnotesize",     where = "!tbp")%
\begin{table}[!tbp]
{\footnotesize
\caption{Trends in metabolism for two sites at each of the NERRS reserves.  Values are averages of mean annual estimates for each period of observation (1994-2004 and 2005-2014). Bold values indicate an increase from the first period, whereas italic values indicate a decrease. Sites were assigned to regions based on approximate geographic coordinates.\label{tab:metab_tab}} 
\begin{center}
\begin{tabular}{lllcllcll}
\hline\hline
\multicolumn{1}{l}{\bfseries Site}&\multicolumn{2}{c}{\bfseries NEM\textsuperscript{a}}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries Pg}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries Rt}\tabularnewline
\cline{2-3} \cline{5-6} \cline{8-9}
\multicolumn{1}{l}{}&\multicolumn{1}{c}{1995-2004}&\multicolumn{1}{c}{2005-2014}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{1995-2004}&\multicolumn{1}{c}{2005-2014}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{1995-2004}&\multicolumn{1}{c}{2005-2014}\tabularnewline
\hline
{\bfseries Atlantic}&&&&&&&&\tabularnewline
~~acebb&-0.04&{\bf -0.03}&&5.11&{\it 3.81}&&-5.15&{\bf -3.82}\tabularnewline
~~acesp&-0.26&{\bf -0.1}&&4.68&{\bf 4.93}&&-4.94&{\it -5.03}\tabularnewline
~~cbmip&-0.14&{\it -0.17}&&1.02&{\bf 1.91}&&-1.16&{\it -2.08}\tabularnewline
~~cbmmc&-0.43&{\bf -0.24}&&6.06&{\bf 10.13}&&-6.48&{\it -10.38}\tabularnewline
~~cbvgi&0.13&{\it 0}&&6.56&{\it 4.47}&&-6.43&{\bf -4.46}\tabularnewline
~~cbvtc&-0.15&{\bf -0.07}&&5.58&{\bf 6.07}&&-5.73&{\it -6.14}\tabularnewline
~~delbl&-0.59&{\bf -0.16}&&6.68&{\it 6.51}&&-7.27&{\bf -6.67}\tabularnewline
~~delsl&-0.37&{\bf -0.33}&&2.84&{\bf 3.22}&&-3.21&{\it -3.55}\tabularnewline
~~grblr&&-0.02&&&1.28&&&-1.31\tabularnewline
~~grbor&&-0.04&&&3.16&&&-3.2\tabularnewline
~~gtmfm&-0.09&{\bf -0.04}&&3.36&{\bf 4.19}&&-3.45&{\it -4.23}\tabularnewline
~~gtmpc&-0.48&{\bf -0.45}&&1.95&{\bf 2.91}&&-2.43&{\it -3.37}\tabularnewline
~~hudsk&0.02&{\it 0.02}&&-0.36&{\it -0.39}&&0.38&{\bf 0.41}\tabularnewline
~~hudtn&-0.03&{\it -0.03}&&3.75&{\it 3.43}&&-3.78&{\bf -3.46}\tabularnewline
~~jacb6&0.05&{\it 0.03}&&3.41&{\bf 3.93}&&-3.36&{\it -3.89}\tabularnewline
~~jacb9&0.01&{\it -0.02}&&1.8&{\bf 3.1}&&-1.79&{\it -3.12}\tabularnewline
~~narnc&-0.64&{\bf -0.53}&&12.64&{\it 11.33}&&-13.25&{\bf -11.86}\tabularnewline
~~narpc&-0.14&{\bf -0.03}&&7.36&{\it 6.3}&&-7.5&{\bf -6.32}\tabularnewline
~~niwcb&-0.46&{\bf -0.28}&&5.5&{\bf 6.1}&&-5.95&{\it -6.38}\tabularnewline
~~niwdc&-0.13&{\it -0.2}&&7.28&{\bf 8.02}&&-7.41&{\it -8.23}\tabularnewline
~~nocec&-0.05&{\it -0.06}&&4.86&{\bf 4.88}&&-4.92&{\it -4.94}\tabularnewline
~~nocrc&-0.19&{\it -0.19}&&5.93&{\it 5.31}&&-6.12&{\bf -5.5}\tabularnewline
~~owcbr&-0.17&{\bf -0.09}&&1.45&{\bf 1.9}&&-1.62&{\it -1.99}\tabularnewline
~~owcwm&-0.36&{\bf -0.19}&&7.02&{\it 5.36}&&-7.38&{\bf -5.54}\tabularnewline
~~saphd&-1.28&{\bf -1}&&1.89&{\bf 3.77}&&-3.17&{\it -4.77}\tabularnewline
~~sapld&-0.16&{\it -0.43}&&2.32&{\bf 3.56}&&-2.49&{\it -3.99}\tabularnewline
~~welht&-0.07&{\bf -0.04}&&0.36&{\bf 0.63}&&-0.43&{\it -0.67}\tabularnewline
~~welin&-1.87&{\it -2.49}&&3.61&{\it 2.68}&&-5.48&{\bf -5.18}\tabularnewline
~~wqbcr&-0.01&{\bf 0.01}&&8.13&{\it 7.41}&&-8.14&{\bf -7.4}\tabularnewline
~~wqbmh&0.02&{\it 0}&&2.82&{\it 2.76}&&-2.79&{\bf -2.75}\tabularnewline
\hline
{\bfseries Gulf}&&&&&&&&\tabularnewline
~~apaeb&-0.35&{\it -0.35}&&4.19&{\bf 4.31}&&-4.54&{\it -4.67}\tabularnewline
~~apaes&-0.71&{\bf -0.67}&&3.35&{\it 3.05}&&-4.06&{\bf -3.73}\tabularnewline
~~gndbc&-1.02&{\bf -0.88}&&3.5&{\bf 4.3}&&-4.51&{\it -5.18}\tabularnewline
~~gndbh&-1.81&{\it -2}&&2.19&{\it 1.77}&&-4&{\bf -3.77}\tabularnewline
~~job09&-0.34&{\bf -0.25}&&5.32&{\bf 5.95}&&-5.66&{\it -6.2}\tabularnewline
~~job10&-0.15&{\it -0.15}&&3.58&{\bf 4.03}&&-3.72&{\it -4.2}\tabularnewline
~~marab&&-0.02&&&3.38&&&-3.4\tabularnewline
~~marce&&-0.01&&&3.62&&&-3.63\tabularnewline
~~rkbfb&-0.28&{\it -0.37}&&5.28&{\bf 6.25}&&-5.56&{\it -6.62}\tabularnewline
~~rkbfu&-0.33&{\it -0.41}&&2.95&{\bf 3.24}&&-3.28&{\it -3.64}\tabularnewline
~~wkbfr&-0.29&{\bf -0.14}&&6.27&{\bf 6.35}&&-6.57&{\bf -6.49}\tabularnewline
~~wkbwb&-0.29&{\bf -0.16}&&6.81&{\it 6.54}&&-7.1&{\bf -6.69}\tabularnewline
\hline
{\bfseries Pacific}&&&&&&&&\tabularnewline
~~elkap&0.03&{\bf 0.11}&&14.7&{\bf 19.84}&&-14.67&{\it -19.74}\tabularnewline
~~elknm&-0.24&{\bf -0.08}&&11.95&{\bf 16.37}&&-12.18&{\it -16.45}\tabularnewline
~~kachd&0.2&{\it -0.66}&&7.31&{\it 3.2}&&-7.12&{\bf -3.86}\tabularnewline
~~kacsd&-0.06&{\it -0.26}&&5.15&{\it 4.2}&&-5.21&{\bf -4.45}\tabularnewline
~~pdbbp&0.01&{\bf 0.08}&&6.29&{\bf 7.31}&&-6.27&{\it -7.23}\tabularnewline
~~pdbby&-0.07&{\bf -0.01}&&8.39&{\it 5.92}&&-8.47&{\bf -5.93}\tabularnewline
~~sfbcc&&-0.24&&&3.56&&&-3.79\tabularnewline
~~sfbgc&&-0.55&&&7.32&&&-7.87\tabularnewline
~~sosva&-0.2&{\bf -0.05}&&6.98&{\bf 7.83}&&-7.18&{\it -7.87}\tabularnewline
~~soswi&-0.28&{\bf -0.08}&&4.41&{\bf 4.74}&&-4.69&{\it -4.82}\tabularnewline
~~tjrbr&-1.33&{\bf -0.17}&&7.72&{\it 7.06}&&-9.05&{\bf -7.22}\tabularnewline
~~tjros&-0.26&{\it -0.34}&&10.18&{\bf 11.61}&&-10.44&{\it -11.96}\tabularnewline
\hline
\end{tabular}\end{center}}

\footnotesize{\textsuperscript{a}NEM: net ecosystem metabolism, Pg: gross production, Rt: total respiration, all values in g O$_2$ m$^{-2}$ d$^{-1}$ as annual averages.}\end{table}

The second and third examples are two interactive web applications \citep{Chang15} created using the \pkg{SWMPr} package that illustrate summaries and comparisons of SWMP data.  The first web application evaluates trends in SWMP data within and between sites using an interactive map (Figure~\ref{fig:swmp_comp}): \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}.  Trends between reserves can be viewed using the map, whereas trends at individual sites can be viewed by clicking on a map location.  Site-level trends are shown below the map with a simple linear regression to show an increase or decrease in values over time.  Trends on the map at each station are plotted as circles that identify the direction and significance of the trend, such that larger points with darker colors indicate a strong trend.  The trend over time is blue for decreasing and red for increasing.  The second application provides graphical summaries of water quality, weather, or nutrient station data at individual stations using the \code{plot\_summary} function: \href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary}.  The output is identical to Figure~\ref{fig:summary_ex} with the addition of drop down menus to select the station, date range, and parameter for plotting. Both of the apps provide previously unavailable utilities to interact with SWMP data using an intuitive online interface.  This format may be more appealing for individuals with limited experience using R, while simultaneously illustrating what is possible using functions from the package. The applications have been used extensively with over 300 hours noted in a single month.

@@

I acknowledge the significant work of NERRS researchers and staff that has allowed access to high-quality monitoring data.  Thanks to Todd O'Brien for the inspiration for the online widgets. Thanks to Mike Murrell and Jim Hagy III for assistance with the ecosystem metabolism functions. Thanks to Jeffrey Hollister for providing useful comments on an earlier draft.  The views expressed in this article are those of the author and do not necessarily reflect the views or policies of the U.S. Environmental Protection Agency.  The use of trade names or products does not constitute endorsement by the US Government.

>>>>

\pkg{SWMPr} functions are categorized by one of three steps in the data workflow: \Bigtxt{retrieving}, \Bigtxt{organizing}, and \Bigtxt{analyzing}.  Functions for retrieving are used to import the data into R as a \code{"swmpr"} object class.  Functions for organizing and analyzing the data provide methods for working with a \code{"swmpr"} object.  The following describes the package structure, beginning with the retrieval functions, a description of the \code{"swmpr"} object returned after retrieval, and, finally, the organizing and analyzing functions.

@@

Retrieval functions to import data directly into R from the CDMO include \code{all\_params}, \code{all\_params\_dtrng}, and \code{single\_param}. Due to rate limitations on the CDMO server, the retrieval functions return a limited number of records with each request.  However, the \pkg{SWMPr} functions use the native CDMO web services iteratively (i.e., within a loop) to obtain all requested records.  Download time can be excessive for longer time series.     
\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # all parameters for a station, most recent
> all_params('hudscwq')
> 
> # get all parameters within a date range
> all_params_dtrng('hudscwq', c('09/10/2012', '02/8/2013'))
> 
> # get single parameter within a date range
> all_params_dtrng('hudscwq', c('09/10/2012', '02/8/2013'), 
+   param = 'do_mgl')
> 
> # single parameter for a station, most recent
> single_param('hudscwq', 'do_mgl')
\end{verbatim}
\end{kframe}
\end{knitrout}

The second approach for data retrieval is to use the \code{import\_local} function to import data into R after downloading from CDMO.  This approach is most appropriate for large, specific data requests. The \code{import\_local} function is designed for data from the zip downloads feature in the advanced query section of the CDMO website. The zip downloads feature can be used to obtain data from multiple stations in one request.  The downloaded data will be in a compressed folder that includes multiple .csv files by year for a given data type (e.g., apacpwq2002.csv, apacpwq2003.csv, apacpnut2002.csv, etc.).  The \code{import\_local} function can be used to import files directly from the zipped folder or after the folder is decompressed.

@@

The \code{aggreswmp} function aggregates parameter data for a \code{"swmpr"} object by set units of time.  This function is most useful for aggregating noisy data to evaluate trends on longer time scales or to simply reduce the size of a dataset.  Data can be aggregated by years, quarters, months, weeks, days, or hours by a user-defined function, which defaults to the mean.  A \code{"swmpr"} object is returned for the aggregated data, although the \code{datetimestamp} vector will be converted to a date object if the aggregation period is a day or longer.  Days are assigned to the date vector if the aggregation period is a week or longer based on the \code{round} method for \code{"IDate"} objects created in the \CRANpkg{data.table} package \citep{Dowle14}.  Additionally, the method of treating \code{NA} values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned, particularly for nutrient data (see the example below).

% table for analysis functions
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Analysis functions available from the \\pkg{SWMPr} package.  Full documentation for each function is in the help file (e.g., execute \\code{?aggreswmp} for individual functions or \\code{help.search(`analyze', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:analyze",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Analysis functions available from the \pkg{SWMPr} package.  Full documentation for each function is in the help file (e.g., execute \code{?aggreswmp} for individual functions or \code{help.search(`analyze', package = `SWMPr')} for all).\label{tab:analyze}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\code{aggreswmp}&Aggregate \code{"swmpr"} objects for different time periods - years, quarters, months,  weeks, days, or hours.  The aggregation function defaults to the mean.\tabularnewline
\code{aggremetab}&Aggregate metabolism data from a \code{"swmpr"} object.  This is primarly used within \code{plot\_metab} but may be useful for simple summaries of daily metabolism data.\tabularnewline
\code{ecometab}&Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method.\tabularnewline
\code{decomp}&Decompose a \code{"swmpr"} time series into trend, seasonal, and residual components.  This is a simple wrapper to \code{decompose} \citep{Kendall83}.  Decomposition of monthly or daily trends is possible.\tabularnewline
\code{decomp\_cj}&Decompose a \code{"swmpr"} time series into grandmean, annual, seasonal, and events components.  This is a simple wrapper to \code{decompTs} in the \CRANpkg{wq} package \citep{Jassby14}.  Only monthly decomposition is possible.\tabularnewline
\code{hist}&Plot a histogram for a \code{"swmpr"} object.\tabularnewline
\code{lines}&Add lines to an existing plot created with \code{plot}.\tabularnewline
\code{map\_reserve}&Create a map of all stations in a reserve using the \CRANpkg{ggmap} package \citep{Kahle13}.\tabularnewline
\code{na.approx}&Linearly interpolate missing data (\code{NA} values) in a \code{"swmpr"} object. The maximum gap size that is interpolated is defined by the arguments.\tabularnewline
\code{overplot}&Plot multiple time series on the same y-axis.\tabularnewline
\code{plot}&Plot a univariate  time series for a \code{"swmpr"} object.  The parameter name must be specified.\tabularnewline
\code{plot\_metab}&Plot ecosystem metabolism estimates after running \code{ecometab} on a combined \code{"swmpr"} object.\tabularnewline
\code{plot\_summary}&Create summary plots of seasonal/annual trends and anomalies for a water a single paramter of interest.\tabularnewline
\code{smoother}&Smooth \code{"swmpr"} objects with a moving window average.  Window size and sides (e.g., centered) can be specified, passed to \code{filter}.\tabularnewline
\hline
\end{tabular}\end{center}

@@

{\centering \includegraphics[width=\textwidth]{figure/smooth_ex-1}

@@

{\centering \includegraphics[width=\textwidth]{figure/interp_ex-1}

@@

{\centering \includegraphics[width=\textwidth]{figure/decomp_ex1-1}

@@

{\centering \includegraphics[width=\textwidth]{figure/decomp_ex2-1}

@@

{\centering \includegraphics[width=\textwidth]{figure/summary_ex-1}

@@

The \code{overplot} function provides an alternative approach to viewing observed data from the same station.  This function uses the base \pkg{graphics} package to plot multiple time series on the same y-axis.

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> ## import data
> data(apacpwq)
> dat <- qaqc(apacpwq)
> 
> ## plot
> overplot(dat, select = c('depth', 'do_mgl', 'ph', 'turb'),
+   subset = c('2013-01-01 0:0', '2013-02-01 0:0'), lwd = 2)
\end{verbatim}
\end{kframe}\begin{figure}[!h]

{\centering \includegraphics[width=\textwidth]{figure/overplot-1}

\caption[The \code{overplot} function plots multiple variables on the same y-axis]{The \code{overplot} function plots multiple variables on the same y-axis.}\label{fig:overplot}
\end{figure}

Estimates of ecosystem metabolism provide a measure of overall system productivity to evaluate whether an ecosystem is a net source or sink of organic material.  The open-water method \citep{Odum56} is a common approach to quantify metabolism using a mass balance equation that describes the change in dissolved oxygen over time from the balance between photosynthetic and respiration processes, corrected using an empirically constrained air-sea gas diffusion model \citep{Ro06,Thebault08}. A detailed discussion of the method is beyond the scope of this article, although users are strongly encouraged to consult references herein for additional information (see \citet{Kemp12,Needoba12,Caffrey14}, also the package help files).  The following is an example that shows use of the \code{ecometab} function from a combined water quality and weather data set.  Monthly aggregations of the raw, daily estimates are plotted using \code{plot\_metab} (Figure~\ref{fig:metab_ex}).

@@

{\centering \includegraphics[width=\textwidth]{figure/metab_ex-1}

@@

{\centering \includegraphics[width=\textwidth]{figure/map_ex-1}

@@

@@

Two \CRANpkg{shiny} web applications hav been created that illustrate the improved ability to synthesize and evaluate multi-year time series using \pkg{SWMPr}.  The first application evaluates trends in SWMP data within and between sites using an interactive \pkg{leaflet} map (\citet{Cheng15}, Figure~\ref{fig:swmp_comp}): \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}.  Trends between reserves can be viewed using the map, whereas trends at individual sites can be viewed by clicking on a map location.  Site-level trends are shown below the map with a simple linear regression to show an increase or decrease in values over time, whereas trends between sites are shown on the map for each station as circles that identify the direction and significance of the trend.  The second application provides graphical summaries of water quality, weather, or nutrient station data at individual stations using the \code{plot\_summary} function: \href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary}.  The output is identical to Figure~\ref{fig:summary_ex} with the addition of drop down menus to select the station, date range, and parameter for plotting.

@@

I acknowledge the significant work of NERRS researchers and staff that has allowed access to high-quality monitoring data.  Thanks to Todd O'Brien for the inspiration for the online widgets. Thanks to Mike Murrell and Jim Hagy III for assistance with the ecosystem metabolism functions. Thanks to Jeffrey Hollister for providing useful comments on an earlier draft.
